### 2009-11-27 docs/marcello-dissertation/thesis/chapter5-taxonomic-persistence-storage-selection-for-netbeams.tex
Major revision of the thesis structure, and review of chapters, moving chapter 7 to the appendix, and adding the first revision of the abstract in the preamble.
Chap 3. Reviewed the structure, misspelled words, sentences, etc;

Chap 4. Changes to the structure of the case study. The documentation regarding netbeams was pushed to the last chapter: appendix;

Chap 5. The selection of the technology follows the review with Prof. Puder. One section describes each of the technologies briefly, and details their characteristics during the analysis of each of the taxonomies;

Chap 6. Describes the imp

This chapter analyzes each of the taxonomies defined in chapter
\ref{chap:taxonomies} against the requirements defined by NetBEAMS as our case
study in chapter \ref{sec:problem-requirements}, following the scope of a given
solution also defined in that chapter. The contenders were drawn from a mixed
list of technologies either related in the literature review of chapter 2 or
from the Internet.

@@

This section describes the database systems contenders considered to be used
for the implementation of Data Persistence layer for NetBEAMS. As described in
the previous chapter, the selection of the technologies must be based on the
list of functional and non-functional requirements specified within the scope
of implementation defined.

One of the primary questions related to data persistence is regarding the data
model and type of the database, such as the use of relational model or
structured models. In addition to that, and given the current time of academic
research and new trends distributed computing such as Cloud Computing and the
Internet \cite{cloud-comp-survey}, the selection of a data persistence for
sensor networks should maintain a wider spectrum of the types of database
systems. Among other intriguing questions, \cite{db-is-rdbs-dommed} reflects 
about the use of alternatives to the relational model for dynamic
environments such as data persistence for the Web. Similarly, these
trends are also related to the advances of Cloud Computing and its persistence
layer as a solution to dynamic systems \cite{cloud-comp-architectures}. For
this reason, this work also included the overview of alternative data models
for the development of an experiment that brings data persistence to NetBEAMS.

\begin{itemize}
  \item \textbf{MySQL}: an open-source relational database system that is the
  primary choice of the development of projects of different sizes. It was selected
  because of its popularity in the current community;
  \item \textbf{Oracle}: is the most popular commercial database system, having
  a campus wide academic licenses availabe;
  \item \textbf{TinyDB}\cite{tinydb}: as referred in the literature review,
  TinyDB is a database system used in many different sensor networks implementation
  \cite{sn-db-tinydb};
  \item \textbf{MongoDB}\cite{mongodb}: an open-source key-value database system
  described in the article that questions the use of the relational model
  \cite{db-is-rdbs-dommed}, as well as the Cloud Computing trends survey
  \cite{cloud-comp-architectures}.
  \item \textbf{DB2}\cite{db2}: in the search of an XML database system, DB2 is
  listed as a hybrid database system that offers both XML and relational model
  \cite{db-xml-enabled}. It was selected because NetBEAMS can represent the
  collected data in the XML format \ref{sec:dsp-message}.
\end{itemize}

The following sections analyzes the taxonomies defined in chapter 3, along with
the requirements for the Data Persistence Layer for NetBEAMS, in section 4.

@@

Considering the infrastructure of NetBEAMS, as described in the previous
chapter, and the characteristics of the sensor network defined at the SF-BEAMS
site \cite{sfbeams2006}, the data characteristics are as follows:

\begin{itemize}
  \item Data is generated by sensor devices, such as the YSI sonde
  \cite{YSI-Sonde}, and manually collected by using a laptop to the network
  sink at the RTC laboratories;
  \item Upon data reception, the RTC staff index and archive the data for
  for distribution using the OPEnDAP format;
  \item Another approach for data collection of data generated by the SF-BEAMS
  is the use of NetBEAMS \cite{netbeams2009}, as described in section
  \ref{netbeams-architecture}.
\end{itemize}

As summarized in the previous section, the nature of data from NetBEAMS through
SF-BEAMS is used for the purpose of \textbf{Data Archival}.

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Oracle}: Supports
  \item \textbf{TinyDB}: Supports
  \item \textbf{MongoDB}: Supports
  \item \textbf{DB2}: Supports
\end{itemize}

\section{Analysis of the Location of Sensor Data Taxonomy}
\label{sec:sn-data-location}

Some observations regarding the location of the collected sensor data for
NetBEAMS were described in the previous section and summarized as follows:

\begin{itemize}
  \item NetBEAMS's architecture is based on a single-hop star infrastructure 
   with a direct network sink being the RTC server \ref{sec:sn-infrastructure}.
\end{itemize}

Based in the observations also described in the previous chapter, the type used
for storing data can be either the \textbf{External Storage}. However, the use
of a \textbf{Data-Centric Storage} approach can be considered. The former is the
current approach used by the SF-BEAMS sensor network, while the latter can be
explored in the experiments.

@@

The External Storage approach is the most common way to implement persistence
for any type of systems, including sensor networks, due to its the simplistic
data management. Furthermore, this type of setup is easier to use of
distributed systems approaches such as Replication to help preventing scaling
the system.

@@

The advantage of the Data-Centric approach lies on the way that data is
distributed into different locations. By using techniques of distributed
database systems, the approach improves overall performance of the use of
massive data sets since the operations of the data are directed to specific
locations. Besides, this approach helps the network managers scale the data
storage just by adding new data nodes as required.

In order to support this strategy, approaches such as data partitioning based
on the entities called Database Sharding
\cite{db-shard02}, or the partitioning based on tables
\cite{db-table-partition}. The former is a technique used in distributed
database systems that uses denormalized data models to improve scalability in
terms of data storage and data location, while the latter defines partitioning
techniques in relational databases. It is a type of database system patitioning
where de-normalized data is spread across different other hosts, or shard
\cite{db-shard01}. Different models of database sharding can be developed
\cite{db-shard-schemas}.

\begin{itemize}
  \item Vertical Partitioning: each feature of the system, usually represented
  by a table in the relational model, is placed in a different server host. For
  example, each sensor in a given network can be divided by its device type;
  \item Range Based Partitioning: after partitioning each feature on a
  different server host, different multiple shards, or similar hosts, can be
  added by using the range of a given key of the entity of a function. For
  example, all sensor types in a geographic location may contain its own shard
  based on its IP address;
  \item Key or Hash Based Partitioning: a given value for a given property of
  an instance of data that is chosen to be used in a hash function, whose
  output determines which shard to use;
  \item Directory Based Partitioning: a simple lookup service is used to
  identify which database shard to use, where the lookup table can be stored
  in a secondary area. Each approach for database sharding used be used
  together on a given function depending on the needs for performance.
\end{itemize}

For example, figure \ref{fig:database-sharding-by-region} is an example of
data storage with shards \cite{db-shard01} related the location of data, where
each shard is assumed to be placed in different server host. Since the data
model is denormalized, the organization of data is entirely placed in each of
these shards. As it is shown, the shards by location can be defined by a shard
key, as described before. The horizontal scalability is given related to
positions, while the vertical by the shards added.

@@

%This data organization architecture is currently being used to manage
%structured data that scales to petabytes or more data. One example of such use
%is on BigTable \cite{bigtable}, which provides support to different Google
%products with requirements on high availability and dynamic data models and
%uses automatic sharding to its data. The BigTable project uses MapReduce
%[ds02], a programming model designed to search huge amounts of data in
%paralell.

@@

The data partitioning is a very restricted technology, mostly available to
advanced users of distributed database systems. The main users of wireless
sensor networks do not have such expertise.

\begin{itemize}
  \item Referential integrity: any cross-shard queries can represent problems 
  since the application layer is responsible to enforce data integrity, like
  verifying constraints of foreign keys;
  \item Rebalancing Shards: If the schema model changes for the shards, the
  strategy of rebalancing must be done. For instance, when one of the shards 
  outgrows more than the others. As a consequence, the database shards must be
  rebalanced by transferring data to new locations whenever necessary. This
  technique is starting to appear as COTS implemetations such as
  \cite{mongodb}.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Oracle}: Supports
  \item \textbf{Berkeley TinyDB}: Supports
  \item \textbf{MongoDB}: Supports
  \item \textbf{CouchDB}: Supports
  \item \textbf{DB2}: Supports
\end{itemize}

@@

As a direct result from the previous section, the use of \textbf{Centralized Query
Processing} is indicated for NetBEAMS, given the fact that NetBEAMS contains
a single network sink, and therefore, a centralized location of the data.

@@

Centralized data management and query processing is simpler than in in-network.
The centralized query processing can be

@@

The creation of the so-called Funneling Affect, since the point-of-traffic is
concentrated in the centralized system. Any data exchange will be using the
same channel of data exchange.

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Oracle}: Supports
  \item \textbf{Berkeley TinyDB}: Supports
  \item \textbf{MongoDB}: Supports
  \item \textbf{CouchDB}: Supports
  \item \textbf{IBM DB2}: Supports
\end{itemize}

@@

One of the most common practices in the area of database system is to use the
relational model to persist data, although the application of the system may
not fit to solve the problem. In fact, the main users of sensor networks may
not hold any expertise in database systems or data modeling, given they come
from different science areas. Taking NetBEAMS as an example, we see a Sensor
Network managed by Marine Biologists without expertise in Data Management,
Modeling systems, and for this reason, one of the requirements of the system is
the use of schema-less approaches. The \textbf{Key-Value-Pair Data Model} or the
\textbf{Document-Oriented Data Model} seems to be the simplest choices of the
models.

@@

Considering the inception of a relational data model \cite{relational-model} and
the use of the YSI Sonde data as the main entity in the system, let figure 
\ref{fig:Relational-Model-Original} represent a prototype of the relational
model, after passing through the process of normalization
\cite{db-normalization}.

@@

Supposing a new type is added into the system, let the refactored
version of the relational model be depicted in figure
\ref{fig:Relational-Model-Addition-Modified}.

@@

\begin{itemize}
  \item Constant data schema changes require constant database normalization
  process, changes to structure, database management, etc;
  \item Some research have shown that the Relational Model does not fit the
  nature of collected data from wired or wireless sensor networks. They usually
  does not support time-series data nor provenance.
  \item Even project aiming at updating the relational sensor database systems
  and SQL clauses have been proposed. 
\end{itemize}

Other projects have been using the XML data models. It falls into the same
category as the Relational Model, since the XML documents \cite{xml}, when in
the context of a database, must be complaint to an XML Schema \cite{xml-schema}. As shown in
section 3, this model uses XPath \cite{xml-xpath} technology for querying
documents, although some hibrid technologies may still use SQL \cite{sql} for
that matter. In this way, an instance of such schema dependent types can be
seen in figure \ref{fig:persistence-example-relational}.

@@

Although this data model have been used in different types of applications,
developers have tried to use the Relational Model to model an application that
could prevent any database change through the use of a technique that relates a
key to a value, called Key-Value data model. With the advent of Internet
applications, the need of such a model that does not need constant refactoring
led developers to propose such approach using the relational model as
documented in technical articles such as \cite{db-kvp-in-relational01} and
\cite{db-kvp-in-relational02}. Based in these articles, consider the relational
model depicted in figure \ref{fig:KVP-on-Relational-Model} as model for the
persistence data for our case study using the key-value strategy. The problem
with it is that key repetition occurs, since the nature of time-series data
requires the use of a timestamp key for each property of the sensor. Therefore,
this strategy is not well-suited to provide persistence data from collected
sensor networks.

@@

\begin{itemize}
  \item entities from the same domains are placed into a bucket;
  \item entities have a set of attributes and relating values
  \item entities with different set of attributes may be contained in the same
  bucket, since there is no schema to govern the bucket items restriction.
\end{itemize}

@@

\begin{itemize}
  \item MySQL: NO SUPPORT
  \item Oracle: NO SUPPORT
  \item Berkeley TinyDB: NO SUPPORT
  \item \textbf{MongoDB}: Supports Both
  \item IBM DB2: NO SUPPORT
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Oracle}: Supports
  \item \textbf{Berkeley TinyDB}: NO Supports
  \item \textbf{MongoDB}: Supports
  \item \textbf{CouchDB}: Supports
  \item \textbf{IBM DB2}: Supports
\end{itemize}

@@

\begin{itemize}
  \item Given the fact of the nature of the project, the technology to be used
  must be \textbf{open-source} \cite{open-source}, that is, no costs involved
  in the adoption of such technology;
  \item A solution for persisting collected sensor data must be not only limited
  to a technology that provides data access, such as SQL, but also by other
  \textbf{data access mechanisms} such as through \textbf{native APIs}. The
  reason is that part of the sensor network data does not have skills in
  database models, but may have skills with programming languages;
  \item Supports hot backup without service interruption.
\end{itemize}

@@

Most of the technologies listed provides access to the data sets through the
use of drivers in different programming and scripting languages such as Java,
Python, Perl, etc.

\begin{itemize}
  \item \textbf{MySQL}: Supports
  \item \textbf{Oracle}: Supports, Not Open-Source
  \item \textbf{Berkeley TinyDB}: Supports, Open-Source
  \item \textbf{MongoDB}: Supports
  \item \textbf{CouchDB}: Supports
  \item \textbf{IBM DB2}: Supports, Open-Source
\end{itemize}

@@

Each of the databases were scored according to its support to the different
types of taxonomies, and randomly selected by using the literature review's
list, as well as the Internet.

>>>>

This chapter analyzes each of the taxonomies defined in chapter
\ref{chap:taxonomies} against different other variables. First, the
requirements defined by NetBEAMS, as our case study in chapter
\ref{sec:problem-requirements}, is be compared. Then, the scope of this work is
compared to the technologies selected based on the readings of the
literature review in persistence in sensor networks and other new trends in
performance. Finally, a table summarizes the contenders strenths and discusses
the selected technology to be integrated with the design of a Data
Persistence component.

\section{Scope of the use of the Techonolgy}

The scope of the use of the database can cover different functional and
non-functional requirements defined in the previous chapter. This work does
not intend to implement an application based on the database system selected,
but provide the foundation of a database management system that can be used to
develop one focused in less refactoring and high scalability and performance.
For this reason, the use of the database system will be restricted to the data
persistence and use through native system or through the use of native
programming languages or API. In this way, the scope of this technology
selection is can be summarized as follows:

\begin{itemize}
  \item Provide data management over the collected sensor data, as a regular
  database system;
  \item Must be able to scale in the face of an increase to the data load of
  the sensor network, as described in chapter 2;
  \item Must be able to use single or multiple servers to improve performance
  and optionally deal with Data-Centric approaches to sensor data partitioning;
  \item Must provide a data model that scales, decreasing the constant data
  schema changes.
\end{itemize}

@@

This section describes the database systems contenders considered to be used
together with for the implementation of the Data Persistence component for
NetBEAMS. As described in the previous chapter, the selection of the
technologies must be based on the list of functional and non-functional
requirements specified within the scope this work, defined in the previous
section.

One of the primary questions related to data persistence is regarding the data
model and type of the database, that is, the use of relational model or
structured models. In addition to that, and given the current time of academic
research and new trends distributed computing such as Cloud Computing and the
Internet \cite{cloud-comp-survey}, the selection of a data persistence for
sensor networks should maintain a wider spectrum of the types of database
systems. Among other intriguing questions, \cite{db-is-rdbs-dommed} reflects
about the use of alternatives to the relational model for dynamic environments
such as data persistence for the Web. Similarly, these trends are also related
to the advances of Cloud Computing and its persistence layer as a solution to
dynamic systems \cite{cloud-comp-architectures}. For this reason, this work
also included the overview of alternative data models for the development of
an experiment that brings data persistence to NetBEAMS. In such a fashion, the
list of the systems considered for evaluation is the following:

\begin{itemize}
  \item \textbf{MySQL}\cite{mysql}: an open-source relational database system
  that is the primary choice of the development of projects of different sizes. It was selected
  because of its popularity in the current community;
  \item \textbf{Oracle}\cite{oracle}: is the most popular commercial database
  system, having a campus wide academic licenses availabe;
  \item \textbf{TinyDB}\cite{tinydb}: as referred in the literature review,
  TinyDB is a database system used in many different sensor networks implementation
  \cite{sn-db-tinydb};
  \item \textbf{MongoDB}\cite{mongodb}: an open-source key-value database system
  described in the article that questions the use of the relational model
  \cite{db-is-rdbs-dommed}, as well as the Cloud Computing trends survey
  \cite{cloud-comp-architectures}.
  \item \textbf{DB2}\cite{db2}: in the search of an XML database system, DB2 is
  listed as a hybrid database system that offers both XML and relational model
  \cite{db-xml-enabled}. It was selected because NetBEAMS can represent the
  collected data in the XML format \ref{sec:dsp-message}.
\end{itemize}

These technologies will be reviewed in contrast to their capabilities to adhire
to the specifications of the taxonomies defined in chapter 3 and the
requirements and the scope for a data persistence of NetBEAMS, as defined in
chapter 4.

@@

The execution of the SF-BEAMS sensor network, together with the NetBEAMS
automated infrastructure, can be summarized as follows:

\begin{itemize}
  \item Data is generated by sensor devices, such as the YSI sonde
  \cite{YSI-Sonde}, and manually collected by using a laptop to the network
  sink at the RTC laboratories;
  \item Upon data reception, the RTC staff index and archive the data for
  for distribution using the OPEnDAP format;
  \item An automated approach to the data collection procedures is done
  using NetBEAMS \cite{netbeams2009}, using software components to extract and
  send the data to the network sink. A complete description of this approach
  is described in section \ref{netbeams-architecture}.
\end{itemize}

As summarized in the previous section, the nature of data from NetBEAMS through
SF-BEAMS is used for the purpose of \textbf{Data Archival}. After data is
collected from the sensors, it needs to be processed and archived for reuse by
different types of users, ranging from researchers, students, and the public
over the Internet.

@@

Basically, all the selected database systems support the purpose of the data.

\section{Analysis of the Location of Sensor Data}
\label{sec:sn-data-location}

As seen in the previous section, NetBEAMS's architecture is based on a
single-hop star infrastructure from the SF-BEAMS sensor network, having all
the data collected from the sensors being directly sent to the network sink 
\ref{sec:sn-infrastructure}.

Based in the observations also described in the previous chapter, the simplest
location to store the collected data is through the use of an \textbf{External
Storage}. In addition to that, the use of a \textbf{Data-Centric Storage}
approach is also considered, in situations where the data load in the
network sink reaches its storage capacity.

The Data-Centric storage separates the location of storage of the data based in
any property of the data such as the value of a given key, etc. Such approach
of data partitioning is provided by database system in two different flavors
\ref{db-partitioning-relational}:

\begin{itemize}
  \item \textbf{Horizontal Partitioning}: the data segmentation is done based
  on the values of the table rows in different groups allocated to different
  datasets physically located in different servers. In this way, all the columns
  defined for a given entity are found on each partition. For instance,
  horizontal partitioning can be defined as a collection five years worth of
  historical sensor data being partitioned into five distinct servers, where
  each server has a single year's worth of sensor data;
  \item \textbf{Vertical Partitioning}: data is partitioned vertically in a way
  that the columns of an entity are divied to different database tables. In
  this way, the particular dataset is broken down to different tables. For
  example, if the database is used to store pictures of a given sensor camera 
  in a BLOB column, it would be the most indicated to the scenario where 
  the images are not constantly referenced when compared to the the textual
  data.
\end{itemize}

@@

The External Storage approach is the most common way to implement persistence
for any type of systems since data management is usually offered by most
database systems out-of-the-box. Furthermore, this approach can be accompained
with the use of Distributed Systems approaches such as Replication to help a
system to scaling in terms of performance.

@@

The advantage of the Data-Centric approach can be directly associated with
database partitioning mechanisms described earlier. Since data is distributed
into different locations, this approach the decrease the size of datasets when
the database server performs a query processing in a database table. For this
reason, this approach helps the database administrators to scale their
database infrastructure and improve the database performance.

The horizontal or vertical partitioning strategies are also referred to as
Database Sharding \cite{db-shard-discussion}, or simply Database partitioning
\cite{db-table-partition} \cite{db-partitioning-relational}. Both former and
the latter are used in regular schema-dependent models or schema-less
models, which uses denormalized data models. There are different approaches to
partition data \cite{db-shard-schemas} \cite{db-partitioning-relational}
\cite{db-partitioning-relational-oracle}.

\begin{itemize}
  \item \textbf{Directory Based Partitioning}: a simple lookup service is used
  to identify which database shard to use, where the lookup table can be stored
  in a secondary area;
  \item \textbf{Vertical Partitioning}: each feature of the system, usually
  represented by a table in the relational model and placed in a different
  server host. For example, each sensor in a given network can be divided by 
  its device type;
  \item \textbf{Key or Hash Based Partitioning}: the value of a given key
  property can be used as the key of the shard, or used as an input to a hash
  function, whose output determines which database shard location the data is
  going to be stored;
  \item \textbf{Range-Based Partitioning}: after partitioning each feature on a
  different server host, different multiple shards, or similar hosts, can be
  added by using the range of a given key of the entity of a function. For
  example, different sensor data can be distributed by geographic origin of
  the data.
\end{itemize}

An example of a Data-Centric storage with different shards
\cite{db-shard-intro} can be see in Figure
\ref{fig:database-sharding-by-region}, where data is segregated by the
location of data. Each shard is assumed to be placed in different server hosts.
Since the data model is denormalized, the organization of data is entirely
placed in each of these shards in a Horizontal partition of the entity. As it is
shown, the shards by location can be defined by a shard key.

@@

@@

The data partitioning is a very restricted technology, mostly available to
advanced users of distributed database systems. Despite being a capability
being offered common-off-the-shelf by some popular database systems, this
strategy requires closer database management. Some of the problems related to
this strategy are regarding its execution \cite{db-shard-discussion}:

\begin{itemize}
  \item \textbf{Rebalancing Shards}: If the schema model changes for the
  shards, the strategy of rebalancing must be done. For instance, when one of
  the shards outgrows more than the others. As a consequence, the database
  shards must be rebalanced by transferring data to new locations whenever
  necessary. This technique is starting to appear as COTS implemetations such
  as mongoDB \cite{mongodb} or MySQL \cite{mysql};
  \item \textbf{Referential integrity}: any cross-shard queries can represent
  problems since the application layer is responsible to enforce data
  integrity. Examples include verifying constraints of foreign keys when the
  partition is done using Vertical table partitioning, considering that data
  collections are organized in separate spaces, and physically placed in
  different shard nodes \cite{db-table-partition}.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports External or Data-Centric Storage approaches.
  When Data-Centric Storage is used, both vertical and horizontal partitions
  \cite{db-partitioning-relational} can be chosen, using different strategies to
  partition the data;
  \item \textbf{Oracle}: Supports External or Data-Centric Storage
  approaches. Only supports vertical partitioning with the definition
  of different types \cite{db-table-partition};
  \cite{db-partitioning-relational-oracle} when using Data-Centric storage;
  \item \textbf{Berkeley TinyDB}: Only supports the External Storage;
  \item \textbf{MongoDB}: Supports External or Data-Centric Storage.
  Using the latter one, only the horizontal partition approach is available
  using with automatic shards \cite{db-mongo-partition};
  \item \textbf{DB2}: Supports External and Data-Centric storage. Only supports
  vertical table partition \cite{db-db2-partition} for the data-centric storage;
\end{itemize}

All the selected database systems supports the External Storage approach.
However, they all differ with the support to Data-Centric Storage support.

@@

As a direct result from the previous section, the use of \textbf{Centralized Query
Processing} is indicated for NetBEAMS, also given the fact that SF-BEAMS data
goes directly to the single network sink, and therefore, a centralized location
of the data.

@@

Centralized data management and query processing is simpler than the in-network
query method. Data is verified by a simple database management system without
the risk of data being unavailable, when the data is spread among the network
nodes.

@@

When the Centralized Query processing is used, problems may occur. Depending
on the size of the datasets and the database technology, performance may be a
concern. For example, this approach creates of the so-called Funneling
Affect \cite{sn-storage04}, since the point-of-traffic is concentrated in the
centralized system.

In order to mitigate the problems originated by this query processing
mechanism, techniques such as Data Replication can be used.

@@

All the database systems technology support a centralized query system.

@@

One of the most common practices in the area of database system is to use the
relational model to persist data, although the application of the system may
not fit to solve the problem. In fact, the main users of sensor networks may
not hold any expertise in database systems or data modeling, given they come
from different science areas. Taking NetBEAMS as an example, we see a Sensor
Network managed by Marine Biologists without expertise in Data Management,
Modeling systems, and for this reason, one of the requirements of the system is
try the use of schema-less approaches. The \textbf{Key-Value-Pair Data Model}
or the \textbf{Document-Oriented Data Model} seems to be the simplest choices of the
models.

@@

Considering the inception of a relational data model \cite{relational-model} and
the use of the YSI Sonde data as the main entity in the system, let Figure 
\ref{fig:Relational-Model-Original} represent a prototype of the relational
model, after passing through the process of normalization
\cite{db-normalization}.

@@

Supposing a new type is added into the system, let the refactored
version of the relational model be depicted in Figure
\ref{fig:Relational-Model-Addition-Modified}. Some observations are considered
in this scenario:

@@

\begin{itemize}
  \item Constant data schema changes require constant database normalization
  process, changes to structure, database management, etc. Schema-Dependent
  databases always require refactoring of the schema to accommodate changes to
  the entites properties;
  \item The approach of Relational Model might does not fit the nature of
  collected data from wired or wireless sensor networks, since they tend to
  change over time \cite{db-is-rdbs-dommed}. They usually does not give support
  time-series data nor provenance well \cite{sn-provenance};
  \item Some research have suggested changes to the SQL clauses to better
  support time-series \cite{sn-db-newop}.
\end{itemize}

Other projects have been using the XML data models for persistence. It falls
into the same category as the Relational Model using XML documents \cite{xml},
being complaint to an XML Schema \cite{xml-schema}. This model uses XPath
\cite{xml-xpath} technology for querying documents, although some hibrid
technologies may still use SQL \cite{sql} for that matter \cite{db2}. In this
way, a materialized version of the data from the defined previous schema can be
seen in Figure \ref{fig:persistence-example-relational}.

@@

Although this data model have been used in different types of applications,
developers have tried to use the Relational Model to model an application that
could prevent any database change through the use of a technique that relates a
key to a value, called Key-Value data model. With the advent of Internet
applications, the need of such a model that does not need constant refactoring
led developers to propose such approach using the relational model as
documented in technical articles such as \cite{db-kvp-in-relational01} and
\cite{db-kvp-in-relational02}. Based in these articles, consider the relational
model depicted in Figure \ref{fig:KVP-on-Relational-Model} as a data model for
persisting data to our case study using the key-value strategy. The problem
with it is that key repetition occurs, since the nature of time-series data
requires the use of a timestamp key for each property of the sensor. Therefore,
this strategy is not well-suited to provide persistence data from collected
sensor networks.

@@

\begin{itemize}
  \item entities from the same domains are placed into a ``bucket'';
  \item entities have a set of attributes and relating values
  \item entities with different set of attributes may be contained in the same
  bucket, since there is no schema to govern the bucket items restriction.
\end{itemize}

@@

Since the selected data model is the schema-less one, the following can be see
from the technologies:

\begin{itemize}
  \item MySQL: Only supports the Relational Model;
  \item Oracle: Only supports Relational Model;
  \item Berkeley TinyDB: Only supports the Relational Model; 
  \item \textbf{MongoDB}: Supports Document-Oriented Model;
  \item IBM DB2: Only Supports the Relational Model.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports Data Replication;
  \item \textbf{Oracle}: Supports Data Replication;
  \item \textbf{Berkeley TinyDB}: NO Supports to distributed data;
  \item \textbf{MongoDB}: Supports Data Replication;
  \item \textbf{IBM DB2}: Supports Data Replication.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{Open-source}: Given the fact of the nature of the project, the
  technology to be used must be free of cost \cite{open-source}, that is, no
  costs involved in the adoption of such technology; Together with it is the
  support from the community;
  \item \textbf{Native APIs}: In order to cover the scope of this work defined
  in the previous chapter, the solution for persisting collected sensor data
  must be not only limited to a technology that provides data access, such as
  SQL, but also by other \textbf{data access mechanisms};
  \item \textbf{Hot Backup}: Supports hot backup with less service interruption. 
\end{itemize}

@@

Most of the technologies listed provides access to the data sets through the
use of drivers in different programming and scripting languages such as Java,
Python, Perl, etc. Moreover, only TinyDB does not support hot backup.

\begin{itemize}
  \item \textbf{MySQL}: Is open-source, supports hot back and contains
  scripting APIs from the community;
  \item \textbf{Oracle}: It is not Open-Source, but contains lots of support
  from the community. Also APIS are available for the majority of languages;
  \item \textbf{Berkeley TinyDB}: Not open-source;
  \item \textbf{MongoDB}: An open-source database system with an increasing
  support from the community, offering native APIS in different programming
  languages and scripting languages;
  \item \textbf{IBM DB2}: Not Open-Source, but offers support from its
  community.
\end{itemize}

Only MySQL and MongoDB supports the non-functional requirements defined in this
section.

@@

Each of the databases were scored according to its support to the different
types of taxonomies, and randomly selected by using the literature review's
list, as well as the Internet. Since mongoDB is based on a schema-less database
system, and supports most of the taxonomies defined in the previous chapter.

### 2009-11-27 docs/marcello-dissertation/thesis/chapter6-data-persistence-component-design-and-architecture.tex
Major revision of the thesis structure, and review of chapters, moving chapter 7 to the appendix, and adding the first revision of the abstract in the preamble.
Chap 3. Reviewed the structure, misspelled words, sentences, etc;

Chap 4. Changes to the structure of the case study. The documentation regarding netbeams was pushed to the last chapter: appendix;

Chap 5. The selection of the technology follows the review with Prof. Puder. One section describes each of the technologies briefly, and details their characteristics during the analysis of each of the taxonomies;

Chap 6. Describes the imp

The deployment of the persistence layer is comprised of two main
components: the DSP Data Persistence and the MongoDB Database Server. Two
different approaches can be taken for the deployment: a single server of a
cluster of mongoDB instances for a sharded environment.

>>>>

The deployment of the persistence layer is comprised by the addition of two
new components to the DSP Platform: the DSP Data Persistence and the MongoDB
Database Server. The former is responsible for capturing any data
produced by sensors in the data sink, as described in the previous sections of
this chapter. On the other hand, the latter is where the data is going to be
stored, as a result of the technology selection on the previous chapter.

The deployment of the DSP Data Persistence follows the specifications of how to
deploy a DSP Component on an OSGi Framework. For the development of this work,
the Knopflerfish OSGi Container was used, as described in section
\ref{sec:dsp-data-persistence-implementation}. In contrast, the
deployment of the mongoDB can be accomplished in different approaches, as
summarized:

\begin{itemize}
  \item \textbf{Single Server}: a regular deployment of a database system as it
  is done in general applications, where data is managed by one single process;
  \item \textbf{Sharded Server}: a deployment approach where data is sharded
  into different servers.
\end{itemize}
